MinCall --- MinION end2end deep learning basecaller

### 2017-06-30 paper/main.tex

\documentclass[runningheads,a4paper]{llncs}
\usepackage{makeidx}
\usepackage{booktabs}
\usepackage{mathtools}
\usepackage{amsmath}
\usepackage{amsfonts}
% \usepackage{amssymb}
% \usepackage{amsthm}
\usepackage[utf8]{inputenc}
\usepackage{acronym}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{float}
\usepackage{graphicx}
\usepackage{cite}

@@

The exact error rate metric is unreliable since multiple pipeline tools could be the issue. First the sample is prepared, hopefully, uncontaminated and matching reference genome as close as possible, then sequenced using the MinION device obtaining raw data. Next, our model (or other groups ones) come along, basecall the sequence. To evaluate error rate metric basecalled read is aligned to the reference genome using aligners with their own errors/biases, mostly commonly used BWA-MEM~\cite{li2013aligning} and Graphmap~\cite{sovic2016fast}.

\section{Background}

@@

\section{Method}

Instead of opting for the traditional path using HMM or newly adopted RNN we tried using CNN~(Convolutional neural networks)~\cite{lecun-98}, that is their residual version~\cite{he2016deep}. For loss, we used CTC (Connectionist temporal classification)~\cite{graves2006connectionist} between basecalled and the target sequence. The implementation used is open source warp-ctc~\cite{warpctc}. Main computation framework is tensorflow~\cite{tensorflow2015-whitepaper}.

\subsection{CTC}
In this subsection we go into more detail outlining CTC loss and how we applied it.

\section{Data preprocessing}

Dataset was obtained from (Ask somebody, Marko doesn't know!). In this research, models were trained on the E. Coli K-12 strand. (Should I cite someone for strand). Data was split into train and test subset, such that aligned reads map to different reference genome parts. For initial model data bootstrapping metrichorn was used, version (Ask Someone?).

@@

\section{Residual arhitecture}

We use gated residual network, that is layer function is like this: $f(x) = k g(x) + x$ Where $x$ is input layer and $k$ constant learnt during training. Specifically, $g$ used is multiple, 1--3 times in our experiments, Relu-BatchNorm-CNN layers.

\section{CTC}

This section gives a brief overview over CTC, for further detail and in depth explanation we recommend original paper~\cite{graves2006connectionist} or searching for contemporary blog posts.

In CTC we have target sequecne $\mathbf{l}$ consisting from symbols of alphabet $\Sigma$. Our models outputs $\mathbf{x}$, and CTC loss maximizes $p(\mathbf{l}|\mathbf{x})$.

$\mathbf{x}$ is consistend of $n$ independent discrete random variables, $X_i$  over domain $\Sigma \cup \text{Blank}$. Path $p$ through $\mathbf{x}$ is assigning for each $X_i$ single value. It's log probabity is $\log P(p) = \sum_i {\log P(X_i=x_i)}$ where $x_i$s define the path.

We further have merging operator on path, $\text{merge}(x_1, x_2, \ldots, x_n)$ which merges together adjacent equivalent elements. For example $\text{merge}(AAAGC) = AGC$.

Then $p(\mathbf{l}|\mathbf{x}) = \sum_{p\in\text{paths}(\mathbf{x}, \text{merged}(p) = \mathbf{l}}{P(p)}$ or less formally, $P(\mathbf{l}|\mathbf{x})$ is sum of all path probabilites which when merged give $\mathbf{l}$

In out concrete $x$ application $\Sigma = \{\text{A G T C A' G' T' A'}\}$

For decoding, we use beam search decoder, with beam width 100.

@@

>>>>

\documentclass[runningheads,a4paper]{llncs}
\usepackage{makeidx}
\usepackage{booktabs}
\usepackage{mathtools}
\usepackage{amsmath}
\usepackage{amsfonts}
% \usepackage{amssymb}
% \usepackage{amsthm}
\usepackage[utf8]{inputenc}
\usepackage{acronym}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{float}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{hyperref}
% \hypersetup{pdftex,colorlinks=true,allcolors=blue

@@

The exact error rate metric is unreliable since multiple pipeline tools could be the issue. First the sample is prepared, hopefully, uncontaminated and matching reference genome as close as possible, then sequenced using the MinION device obtaining raw data. Next, our model (or other groups ones) come along, basecall the sequence. To evaluate error rate metric basecalled read is aligned to the reference genome using aligners with their own errors/biases, mostly commonly used BWA-MEM~\cite{li2013aligning} and Graphmap~\cite{sovic2016fast}.
\section{Background}

@@

\section{Method}
Instead of opting for the traditional path using HMM or newly adopted RNN we tried using CNN~(Convolutional neural networks)~\cite{lecun-98}, that is their residual version~\cite{he2016deep}. For loss, we used CTC (Connectionist temporal classification)~\cite{graves2006connectionist} between basecalled and the target sequence. The implementation used is open source warp-ctc~\cite{warpctc}. Main computation framework is tensorflow~\cite{tensorflow2015-whitepaper}.

\section{Data}

Dataset used were E.Coli K-12 strands from~\cite{loman1-100k} and Lamba basecalling\footnote{Acquired from doc. dr. sc. Petra Korać i dr.sc. Paula Dobrinić}. In this research, models were trained on the E. Coli K-12 strand. (Should I cite someone for strand). Data was split into train and test subset, such that aligned reads map to different reference genome parts. For initial model data bootstrapping metrichorn was used, version (Ask Someone?).

@@

@@

Also some data were obtained in cooperation with Biologists doc. dr. sc. Petra Korać and dr.sc. Paula Dobrinić. Other were procured publicly shared from Loman labs~\cite{loman1-100k}
